---
title: "Statistics with R"
author: 
  - Professor Stefano Peluso #stefano.peluso@unicatt.it
  - Notes by Maria Briones
output: 
    rmdformats::readthedown
---
```{r Setup, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}

#function to make colored letters
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}

#Set course working directory
mypath <- c("C://Users//marub//Desktop//DataSci for Management//Statistics with R//data")

library(pacman)

p_load(dplyr)
```
 
<!-- 02-02-2026  -->

The topics we will see on this course are under the category of **inferential statistics**: (1) Population and sample, (2) Point estimation, (3) Confidence intervals, (4) Hypothesis testing and (5) Linear regression. We will look at this topics theoretically and apllied in R.
 
# Distributions

## Normal distribution

On the Preparatory Course, we did some plots related to random variables. Now we'll look at the density of a standard Gaussian. We must establish the number of points, a sequence for the domain of the function and it's image.

We can do the full plot or focus our attention in just one part of the plot

```{r}
#we specify the number of plot points
N = 100

#create a vector of N elements from -5 to 5 (domain)
x = seq(-5,5, length=N) 

#for the y axis (image)
fz = dnorm(x)

par(mfrow=c(1,2)) 
#we plot:
plot(x, fz, #the data 
     type="l", 
     main = "Normal 0,1",
     xlab = "random variable X", 
     ylab = "density", 
     lwd=2 # defines the width of the line
     )
plot(x, fz, 
     type="l", 
     main = "Section of Normal 0,1",
     xlab = "random variable X", 
     ylab = "density", 
     xlim = c(-4.5,-0.15), #we can focus our attention on only one part of the plot
     lwd=2 
     )
```
We can also perform the plot in a more direct way combining functions to plot a $X\thicksim N(0,1)$ distribution and a $X\thicksim N(3,5^2)$:

```{r}
par(mfrow=c(1,2)) 
plot(function(y)dnorm(y),#we create an annonymous function only inside of the plot
     main = "Density of Gaussian (0,1)",
     from=-5, to=5)
plot(function(y)dnorm(y,3,5), 
     from = -10, to=20, 
     main = "Density of Gaussian (3,25)")
```
In order to visualize the effect of an increase on parameters we can plot all function on a same grid:
```{r}
# first N(0,1)
plot(function(y)dnorm(y), 
     from = -20, to=20, 
     main = "density of Gaussian 0,1", 
     col=1 #color picked by number
     )
plot(function(y)dnorm(y, 0, 9), 
     from = -20, to=20, 
     main = "Density of Gaussians", 
     col=2, #color picked by number
     add=T)
plot(function(y)dnorm(y, 12, 1), 
     from = -20, to=20,  
     col=3, #color picked by number
     add=T)
```

With the help of a for loop we can also see how a histogram approximate to a normal density as we increase the sample size.

```{r}
values = c(10, 100, 1000, 10000) #we define sample sizes

par(mfrow=c(2,2))

for (n in values) {
  x = rnorm(n) # generate a n-random sample
  hist(x, 50, col=log(n,10)+1, freq=FALSE, 
       main=paste("Histogram for size",n))
  #we use the lines function to impose a N(0,1) on the graph
  lines(sort(x),
        dnorm(sort(x)),
        col=1, lwd=2) 
}
```

## Uniform distribution 

We can replicate this exercise for the case of a uniform distribution $Y\thicksim U(3,7)$ using the sames values than before (plus 100.000).

```{r}
values = c(100, 1000, 10000, 100000)

# Define parameters
a <- 3  # Min value
b <- 7  # Max value

par(mfrow=c(2,2))

# Create a loop
for (n in values) {
  fy = runif(n, a, b) # generate a n-random sample
  hist(fy, 50, col=log(n,10)+1, freq=FALSE, 
       main=paste("Histogram for size",n))
  #we use the lines function to impose a N(0,1) on the graph
  lines(sort(fy),
        dunif(sort(fy), a, b),
        col=1, lwd=2) 
}

```

### User defined functions

On R there are built-in functions like `dnorm(x)`. But we can also create new functions. for example we can create a function called MyFnc with 3 inputs and an output called out, which performs some operations in the body {}.

```{r eval=FALSE}
#basic structure of a function
MyFnc = function(arg1,arg2,arg3){
  #statement1
  
  #statement2
  
  #statement3
  
  return(out) #we tell the output of the function
}
```

## Lévy distribution

Lévy distribution, is a continuous probability distribution for a non-negative random variable. It is a special case of the inverse-gamma distribution and a stable distribution. Is primarily used to model processes characterized by heavy tails, non-negative values, and frequent, significant short-term changes (non-Gaussian, anomalous diffusion)

Common uses include modeling financial market fluctuations, animal foraging patterns (Lévy flights), network traffic, and physical phenomena like turbulence or particle transport.

There isn't a built-in function on R for this distribution, so we can download a specialized library or build the function ourselves.

```{r Levy}
#we define the function
dlevy = function(x,th){
  #replicate the given math expression 
  f = sqrt(th/(2*pi))*exp(-th/(2*x))*x^(-3/2)
  f = f*(x>0) #to impose the condition
  
  return(f)
}
```

Now `dlevy()` is a function in our environment with two inputs. We can evaluate and create a plot to see the Lévy PDF.

Now that we have created our `dlevy(x,th)` function, we can plt the Levy pdf 
```{r Levy}
par(mfrow=c(1,1))

x = seq(0.01,10, by=0.01)
y = dlevy(x,2)

plot(x,y,
     type="l")
```

We can also redefine the `dlevy()` function with default input values to use similarly to `dnorm`:
```{r}
#we define the function
dlevy = function(x=0,th=1){
  #replicate the given math expression 
  f = sqrt(th/(2*pi))*exp(-th/(2*x))*x^(-3/2)
  
  #impose the condition x>0 because if the logal operator x>0 is false -> f=0
  f = f*(x>0) 
  
  return(f)
}

#we evaluate the function on x=5 ,th=1
dlevy(5)
```

## Laplace distribution

Laplace distribution is a continuous probability distribution named after Pierre-Simon Laplace. It is also sometimes called the double exponential distribution, because it can be thought of as two exponential distributions (with an additional location parameter) spliced together along the x-axis
 
The Laplace distribution (double exponential) is commonly used to model data with higher peaks and fatter tails than a normal distribution, particularly when data is clustered around a center with occasional extreme outliers. It is widely applied in finance for modeling risk, signal processing for noise reduction, hydrology for extreme events, and in machine learning to impose sparsity (Lasso regularization).

As before, we need to create a function for a Laplace density: 
```{r}
#definition of the function:
dlaplace = function(x,th){
  #th will be a vector containing both arguments
  mu = th[1]
  b = th[2]
  
  #replicate the given math expression
  f = 1/(2*b)*exp(-abs(x-mu)/b)
  
  #impose the condition x>0 because if the logal operator b>0 is false -> f=0
  f = f*(b>0)
  
  return(f)
}

#we evaluate on x=5 , mu=0 , b=1
dlaplace(5, c(0,1))
```

# Central Limit Theorem (CLT)
<!-- 05-02-2026 -->

We know if we have $n$ random variables $X$ with unknown distribution and $E(X)=\mu$, $V(X)=\sigma^2$, ($n>50$) then the sum of each RV: 

$$\sum_{i=1}^{n}{X_i}\approx N(n\mu,n\sigma^2)$$
We can try to see how the CLT works on different distributions and sample sizes. First Let's assume we have a random variable $X\thicksim U(0,1)$ then, we know by the uniform properties that $E(X)=1/2$ and $V(X)=1/12$. 

```{r}
# sample size 
n=2

# Number of simulations
M = 10000

#generate the rvs
X = runif(M*n, #number of observations
          min=0, max=1)
```

Now we generate a matrix were each column from 1 to $n$ is a random variable with uniform distribution (0,1) and the rows are the $M$ extractions for each one of those variables. 

```{r}
# X as a matrix
XMat = matrix(X,   #fill the matrix with the values of X
              nr=M, #Number of rows
              nc=n #number of columns
              ) #Note: we do not care about the other on wich the matrix is filled
```

We sum each row, this way we will have $m=1000$ random variables that should have a normal distribution according to the CLT:

$$Y=\sum_{i=1}^{n=2}{X_i}\approx N \left( \frac{n}{2}, \frac{n}{12} \right)$$

```{r}
#sum row by row
Y = rowSums(XMat)
#Y = apply(Xmat,1,sum) #equivalent function

#histogram of sum
hist(Y, 50, col=4, prob=TRUE, 
     main="Histogram of Y~N(1,1/6)")

#we add a normal with the same parameters to compare.
plot(function(x)dnorm(x, n/2, sqrt(n/12)),
     add=TRUE, col=2, lwd=3, xlim=range(Y)) 
```
We can see that, in the case of a uniform distribution, even with a sample of n=2, the normal distribution is a good approximation for the distribution of it's sum (and mean).

## Different levels of n

Of course we can repear the exercise to understand how the distribution of the sample mean evolves when we increase the size of the sample.

```{r, , fig.width=10, fig.height=8}
par(mfrow=c(2,2))
M=10000 # Number of simulations

v = c(1,2,5,10) # sample sizes to sum

#loop 
for(n in v){
  #generate the rvs
  X = runif(M*n)

  # X as a matrix
  XMat = matrix(X,   #fill the matrix with the values of X
               nr=M, #Number of rows
               nc=n #number of columns
               ) #Note: we do not care about the other on wich the matrix is filled

  #sum row by row
  Y = rowSums(XMat)
  #Y = apply(Xmat,1,sum)

  #histogram of sum
  hist(Y, 50, col=(which(v %in% n)+2), prob=TRUE, ylim=c(0,1.5), 
       main= paste("Histogram of Y=Sum{X} with n=", n)
       )

  plot(function(x)dnorm(x, n/2, sqrt(n/12)),
       add=TRUE, col=2, lwd=3, xlim=range(Y)) 
}
```

## Sample variance 

As we know for a generic population the sample variance is given by 

$$s^2=\frac{\sum(X_i-\bar X)^2}{n-1}$$ 
and thus, when $ X \thicksim N(\mu, \sigma ^2)$ then:

$$\frac{(n-1)s^2}{\sigma^2}\thicksim \chi^2_{n-1}$$
Now, let's say that we have the probability $Pr(s^2>3)$ where $s^2$ is the sample variance, when $X \thicksim N(3,4)$ and the sample size of our realization is $n=7$

$$\frac{(7-1)s^2}{4} \thicksim  \chi^2_{7-1}$$
$$ \begin{split} 
  Pr(s^2>3) &= 1-Pr(s^2<3)\\
            &= 1-Pr\left(\frac{6s^2}{4}<\frac{6\cdot 3}{4}\right)
   \end{split}$$ 

which we can compute as:
```{r}
#exact value of the pdf of chi square
1 - pchisq(18/4, 6)
```

If we want to evaluate $Pr(s^2>3)$ without using $\chi^2$, when $X \thicksim N (3,4)$ and $n=7$ we can use a *Montecarlo Procedure*
```{r}
# First we set our parameters: 

M = 10000 # number of simulations

# ---- population parameters 
Mu =3
Sigma = 2
n = 7

#Example of code for montecarlo (innefficient):

#counter of number of times that condition s^2>3 is TRUE
C = 0

for (i in 1:M) {
  #extract a sample
  x = rnorm(n, Mu, Sigma)
  
  #compute sample variance (the function in r uses the sample variance formula)
  s2 = var(x)
  
  #we update our counter if the condition is true
  if(s2>3){
    C = C+1
  }
}

#finally we compute the counter/simulation ratio
C/M

```

An alternative to get the same result 
```{r}

#replacement of the counter
C = rep(0,M)

#we will fill it by 1 when S^2>3
for (i in 1:M) {
  #extract a sample
  x = rnorm(n, Mu, Sigma)
  
  #compute sample variance (the function in r uses the sample variance formula)
  s2 = var(x)
  
  #we update our vector if the contidion is true
  if(s2>3){
    C[i]=1
  }
}

#the advantage is that we can see the evolution of C
plot(cumsum(C)/(1:M)) #cumulative mean
abline(h=1-pchisq(18/4, 6), col=2, lwd=3)
```

This plot can be used as a guidance for fixing M once it stabilizes. For example, with the graph we can see that we do not need 10.000 simulations, but only 2.000. 

A third approach would be 
```{r}
#we create a vector to be filled
Z = c()

#we will fill it by 1 when S^2>3
for (i in 1:M) {
  #extract a sample
  x = rnorm(n, Mu, Sigma)
  
  #compute sample variance (the function in r uses the sample variance formula)
  s2 = var(x)
  
  #we update our vector if the contidion is true
  if(s2>3){
    Z = c(Z,1)
  }else{
    Z = c(Z,0)
  }
}
```

Let's create a function for computing the pooled sample variance 
```{r}
#for two RV: x and y
poolvar = function(x,y){
  #retieve the sample sizes
  n=length(x)
  m=length(y)
  
  #compute the weight of x relative to y 
  alpha=(n-1)/(n+m-2)
  
  #pooled sample variance
  S2 = alpha*var(x)+(1-alpha)*var(y) 
  
  return(S2)
}

sample1 = rnorm(100,0,1)
sample2 = rnorm(50,0,2) #note less size and higher variance


poolvar(sample1, sample2)

```

we can get different results but we must always have TRUE on the following condition
```{r}
var(sample1) < poolvar(sample1, sample2) & poolvar(sample1, sample2) < var(sample2)
```

# Maximum Likelihood Estimator (MLE)

We retrieve the ML estimator for the intermediate case where $(x_1,x_2,x_3)=(1,0,1)$ we had already estimated that $\hat p=2/3)0,66$
```{r}
#we plot the function of the Likelihood 
plot(function(p)p^2*(1-p))
abline(v=2/3, col=2)
```

Now we want to program a user defined function to compute the following matemathical function: 

$$F(x)=  \begin{cases} x^2+2x+3 & x < 0 \\
                     x+3      & 0 \leq x < 2 \\
                     x^2+4x-3 & x \ge 2  \end{cases}$$

```{r}
#create the function
exone = function(x){
  fx = ifelse(x < 0,
              x^2 + 2*x + 3,
              ifelse(x < 2,
                     x + 3,
                     x^2 + 4*x - 7))
  return(fx)
}

#we evaluate
plot(function(x)exone(x), from=-3, to=6)
```
qn## Function optimum

`optimum` is the function we use for numerical optimization. 
Inputs: 

1. Initial value (arbitrary but reasonable)
2. function to maximize
3. min / max 
4. other inputs of the function

We aim to find the mode of a gamma distribution (the point x* at which pdf/pmf is max). Note: A gamma distribution is useful for positive, continuous rvs.

```{r}
#compute the function oprtim por a gamma distribution
optim(1, #initial value 
      dgamma, #function to maximize 
      control=list(fnscale=-1), #with this we maximize (since optim usually minimizes)
      shape=2, rate=2 #parameters needed by function dgamma
      )
```
The output we obtain is a list, **par** is the maximum or minimum point we were looking for (x\* ), **value** is the oprtimized value of f, that is f(x\* ), **counts** is the number of iterations to convergence, **gradient** depends on the algorith we choose, but reflects the approximation of the first derivative. Finally, **convergence** is a sort of check, if it's 0 everything was ok with the process and we get no **message**. 

In sum, the mode is $0.5$ and we know that the exact mode for a gamma distribution is given by: 

$$\frac{shape-1}{rate}=\frac{2-1}{2}=0.5$$

In this example we have a comparison with the analytical solution, but some other times we do not have this comparison. 

Now, we want to find the optimum point for the function 

$$f(x,y)=x-1+\frac{3.2}{y}+ln[\Gamma(x)]+3x*ln(y)$$
As there is not such function in R we must create it 
```{r}
#create the function
f1 = function(x,y){
  z = x-1+(3.2/y)+log(gamma(x))+3*x*log(y)
  return(z)
}

#create an x axis 
x = seq(0.1,
        2, 
        by= 0.1)

#create y-axis
y = x

#evaluate the f1 at each possible couples of x and y (z-axis)
z = outer(x,y,f1)

persp(x, y, z,
      ticktype = "detailed",
      theta=65 #rotate the plot
      )
```
As such `f1` is defined in a useful manner to plot, but not for optim since `optim` uses only the first argument of the function. 

Then, let's modify f1. 
```{r}
f2 <- function(theta){
  x = theta[1]
  y = theta[2]
  
  z = x-1+(3.2/y)+log(gamma(x))+3*x*log(y)
  return(z)
}
```

now we can use optim to minimize our function: 
```{r}
optim(c(1,1), 
      f2)
```
## Normal Likelihood and logLikelihood

Now we create a function for likelihood in the gaussian case $N(\mu, \sigma^2)$. We want the output to be $L(\mu, \sigma, sample)$

```{r include=FALSE, echo=FALSE, eval=FALSE}
#function for log-likelihood
Nlk = function(theta, n){
  mu=theta[1]
  sigma=theta[2]
  
  sample=rnorm(n, mu, sigma)
  
  lnL=-(n/2)*log(2*pi)-(1/2)*sum(sample-mu)^2
  
  return(lnL)
}

Nlk(c(1,2), 5)
```

```{r}
#function definition
NormLik = function(mu,sigma,x){
  fxi <- dnorm(x, mu, sigma) #densities
  L <- prod(fxi) #likelihood
  return(L)
}

#sample 
collected.data=rnorm(5,1,2)

#evaluate 
c(NormLik(1,2,collected.data),
  NormLik(10,2,collected.data) #smaller 
)
```
But, if we try it on a larger sample:
```{r}
#size 1000
collected.data2=rnorm(1000,1,2)

#evaluate
c(NormLik(1,2,collected.data2),
  NormLik(10,2,collected.data2)
)
```
As the results tend to 0 we are not able to discriminate between different parameter values. To adress this issue we turn then to log-likelihood
```{r}
#function definition
NormLogLik = function(mu,sigma,x){
  fxi <- dnorm(x, mu, sigma) #densities
  lfxi <- log(fxi) #log-densities
  logL <- sum(lfxi)
  return(logL)  
}

#evaluate
c(NormLogLik(1,2,collected.data2),
  NormLogLik(10,2,collected.data2))
```
Thanks to this process we are able to discriminate between different parameter values, and $\mu=1$ has much higher loglik. 

Yet, to be able to optimize the parameters we need to aggregate them into a unique vector. 
```{r}
#function definition
NormLogLik2 = function(theta,x){
  mu=theta[1]
  sigma=theta[2]
  
  fxi <- dnorm(x, mu, sigma) #densities
  lfxi <- log(fxi) #log-densities
  logL <- sum(lfxi)
  return(logL)  
}

#eval
c(NormLogLik2(c(1,2),collected.data2),
  NormLogLik2(c(10,2),collected.data2)
)
```

## MLE Exponential

Let's write a function for exponential loglik and then plot the loglik from 0 to 2, for the given sample $x=(1,2,7,2,3)$
```{r}
ExpLogLik = function(lam, x){
  fxi <- dexp(x, lam) #densities
  lfxi <- log(fxi) #log-densities
  logL <- sum(lfxi)
  return(logL)  
}

values = seq(0,
        2, 
        by= 0.01)
x=c(1,2,7,2,3) 

logexp = c()

for (i in 1:length(values)) {
  logexp[i] = ExpLogLik(values[i],x) #log-lik evaluated at the i-element of lams
  }

plot(values, logexp, type="l")

```
If we look for the MLE of $lambda$, note that in this context `optim` has the additional inputs
```{r}
sampled.data=rexp(10,0.2)
MLE <- optim(0.1, 
      ExpLogLik, 
      control=list(fnscale=-1),
      x=sampled.data)
lam.mun <- MLE$par
lam.mun
lamhat = 1/mean(sampled.data)
lamhat
```
## MLE for Normal 

Now let's find the MLE for $\mu$ and sigma in a Gaussian population, then compare numerical and analytically solution on `xvec = rnorm(1000,10,2)`

```{r}
#we already have the NormLogLik2 function, we optimize to find the MLE for mu 
xvec=rnorm(1000,10,2)

MLE2 = optim(c(0,1), 
      NormLogLik2, 
      control=list(fnscale=-1),
      x=xvec)
theta.mle <- MLE2$par
theta.mle

muhat = mean(xvec)
muhat

sigmahat=sqrt(((1000-1)/1000)*var(xvec))
sigmahat
```

Next time: couple of final examples on MLE with a real data vector and on a second example we will see MLE for a used defined pdf.  

```{r clean1, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
rm(list = ls()[ls() != "mypath" & ls() != "colorize"])
gc()
```

## MLE on Gamma

We will find estimates of $\alpha$ and $\beta$ for $x \thicksim \Gamma(\alpha,\beta)$, where $X$ is the closing prince of Apple stock. 
```{r}
# Read de dataframe we will use from the folder data on my working directory
data = read.csv(paste0(mypath,"//Apple.csv")
                          , sep=",")

#store on environment the closing price
x = data$Close

#define a function to obtain the log-likelihood for a gamma function. 
Gll = function(theta, x){
  #define pdf
  f = dgamma(x, shape=theta[1], scale=theta[2]) #we can choose rate OR scale
  
  #log-likelihood
  ll = sum(log(f))
  
  return(ll)
}

MLE.gamma <- optim(c(1,1),
                   Gll, 
                   control=list(fnscale=-1),
                   x=x)

MLE.gamma$par
```

Then, if we want to compute the probability that the price is below 550
```{r}
pgamma(550, MLE.gamma$par[1], scale=MLE.gamma$par[2])
```
Now we wish to find the MLE of pars of Kumaraswami distribution, a generalization of the unifor, with two parameters a and b. 
It reduces to the uniform when a=b=1. We need to do the following steps: 

1. function for pdf 
2. definition for log-likelihood
3. oprtimization for ll

```{r kumaraswaniMLE}
#TO DO: define errors when parameters do not met conditions
dKuma <- function(x,a,b){
  #definition of the PDF
  fx =a*b*(x^(a-1))*((1-(x)^a))^(b-1)
}

Kll = function(theta, x){
  #define pdf
  f = dKuma(x, a=theta[1], b=theta[2]) #we can choose rate OR scale
  
  #log-likelihood
  ll = sum(log(f))
  
  return(ll)
}

#define the sample
unifsample = runif(100, 0, 1)

MLE.Kuma = optim(c(0.01,0.01),
                   Kll, 
                   control=list(fnscale=-1),
                   x=unifsample)

MLE.Kuma$par

```
values for a, and b close to one make sence since obs has been generated from an uniform and we know that kumawasaki tends to a uniform when a=b=1. 

# Method of Moments

Let's find the MM esttimates for a Binomial distribution 
```{r}
set.seed(12345)

#sample size 
n=4

#Binomial paramethers
N = 10
p = 0.1

#Binomial random sample 
x=rbinom(n,N,p)

#MM estimates 
xbar = mean(x)
N.tilde = xbar^2/(xbar-sum((x-xbar)^2)/n)
p.tilde = xbar/N.tilde

theta = c(xbar, N.tilde, p.tilde)
theta
```

now lets try with increasing n
```{r}
ns=seq(4,10000,by=10) #samples sizes we will try 

#empty vector to fill
p.vec = c()
N.vec = c()
    
for (n in ns) {
  x = rbinom(n,N,p)
  
  #compute MM estimates 
  xbar = mean(x)
  N.tilde = xbar^2/(xbar-sum((x-xbar)^2)/n)
  p.tilde = xbar/N.tilde
   
  #store the estimates
  p.vec = c(p.vec, p.tilde)
  N.vec = c(N.vec, N.tilde)
}

par(mfrow=c(1,2))

#plot for p
plot(ns, p.vec)
abline(h=p, 
       col=2, 
       lwd=3)
#plot for n
plot(ns, N.vec)
abline(h=N, 
       col=2, 
       lwd=3)
```


```{r clean2, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
rm(list = ls()[ls() != "mypath" & ls() != "colorize"])
gc()
```

# Confidence Intervals

## Normal distribution and known variance

We are in need of a function to compute $CI_(1-\alpha)(\mu)$ for a random variable $X \thicksim N(\mu, \sigma ^2)$ where $\sigma$ is known. 

As inputs then we have the sampled data, the level of confidence, the population variance and we need to estimate the confidence interval for the unknown parameter.

```{r}
#customized function
Normal_CI <- function(x, lvl=0.95, sigma=1){
  
  #known parameters
  alpha = 1-lvl
  n = length(x)
  
  #---------list to be filled
  params <- list(1,2,3,4,5)
  param_names <- c("Computed values",
                   "Lower bound", "Upper bound",
                   paste0("Confidence Interval at ",lvl*100,"%"),
                   "CI width")
  names(params) <- param_names
  
  #computed values
  z = qnorm(1-alpha/2)
  MoE = z*sigma/sqrt(n) #margin of error
  avg = mean(x)
    
  comp = c(avg, sigma, z,MoE)
  names(comp) = c("Sample Mean", "Population SD", "Z-quantile", "Margin of Error")
  params[[1]] <- comp
  
  #lower bound
  L1 = avg-MoE
  params[[2]] <- L1
  
  #upper bound
  L2 = avg+MoE
  params[[3]] <- L2
  
  #Confidence intercal 
  CI = c(L1, L2)
  params[[4]] <- CI
  
  #width
  w <- diff(CI)
  params[[5]] <- w
  
  return(params)
}

#generate a random sample
x = rnorm(16, 25, sd=sqrt(36))

#eval
a <- Normal_CI(x, sigma=sqrt(36))
a
```
We can check how the CI changes as we modify our level of confidence: 
```{r}
#eval
b <- Normal_CI(x, lvl=0.99, sigma=sqrt(36))
b
```
## Normal distribution and known variance

As we lost certanty we won't be able to use $z$ (normal distribution) anymore, instead we must use $t$. 
```{r}
#customized function
unkNormal_CI <- function(x, lvl=0.95){
  
  #known parameters
  alpha = 1-lvl
  n = length(x)
  
  #---------list to be filled
  params <- list(1,2,3,4,5)
  param_names <- c("Computed values",
                   "Lower bound", "Upper bound",
                   paste0("Confidence Interval at ",lvl*100,"%"),
                   "CI width")
  names(params) <- param_names
  
  #computed values
  avg = mean(x)
  s = sd(x)
  T = qt(1-alpha/2, n-1)
  MoE = T*s/sqrt(n) #margin of error

  comp = c(avg, s, T, MoE)
  names(comp) = c("Sample Mean", "Sample SD", "T-quantile", "Margin of Error")
  params[[1]] <- comp
  
  #lower bound
  L1 = avg-MoE
  params[[2]] <- L1
  
  #upper bound
  L2 = avg+MoE
  params[[3]] <- L2
  
  #Confidence intercal 
  CI = c(L1, L2)
  params[[4]] <- CI
  
  #width
  w <- diff(CI)
  params[[5]] <- w
  
  return(params)
}
```

For our previous sample we know that $CI_{95\%}(\mu)=[21.03983, 27.84352]$ with a CI width of $5.879892$ for this same sample as we "don't know" the population variance anymore we expect a wider CI.

```{r}
#eval
unkNormal_CI(x)
```
There is actually a build-in function on r to perform this calculus: 
```{r}
t.test(x, conf.level = 0.95)
```
We can use the `t.test()` function in any case were we have normal distributions and unknown parameters.

## Bernoulli asymptotic CI 

A CI for large samples on the case of a Bernoulli. On this case the input we have are the sumber of sucesses, the sample size and the confidence level. 
```{r}
#customized function
Ber_AsyCI <- function(suc, n, lvl=0.95){
  
  #known parameters
  alpha = 1-lvl

  #---------list to be filled
  params <- list(1,2,3,4,5)
  param_names <- c("Computed values",
                   "Lower bound", "Upper bound",
                   paste0("Confidence Interval at ",lvl*100,"%"),
                   "CI width")
  names(params) <- param_names
  
  #computed values
  prop = suc/n
  var = prop*(1-prop)
  z = qnorm(1-alpha/2)
  MoE = z*sqrt(var)/sqrt(n) #margin of error

  comp = c(prop, sqrt(var), z, MoE)
  names(comp) = c("Sample Mean", "Sample SD", "z-quantile", "Margin of Error")
  params[[1]] <- comp
  
  #lower bound
  L1 = prop-MoE
  params[[2]] <- L1
  
  #upper bound
  L2 = prop+MoE
  params[[3]] <- L2
  
  #Confidence intercal 
  CI = c(L1, L2)
  params[[4]] <- CI
  
  #width
  w <- diff(CI)
  params[[5]] <- w
  
  return(params)
}

#generate random sample
ber <- rbinom(344, 1, 0.24)

#eval 
Ber_AsyCI(sum(ber), 344, lvl=0.9)
```
## Normal asymptotic CI 

As we stated before, on the case on unkwnown parameters we can use the `t.test()` function. Let's explore this aproximation with the *Apple.csv* dataset.
```{r}
data = read.csv(paste0(mypath,"//Apple.csv")
                          , sep=",")
x = data$Open
y = data$Close

#t test with two samples
t.test(x, y, conf.level = 0.95,
       var.equal = TRUE #if we assume equal variances
       ) 
```
As we can see the confidence interval includes both negative and positive values, so we can conclude that there is not evidence to state that *Open* is higher than *Close* or viceversa. Also, as the length of *data* es $78$ we can notice that $df=154 = 78+78-2$.

```{r}
#t test with two samples
t.test(x, y, conf.level = 0.95,
       var.equal = FALSE
       ) 
```
Here the degrees or freedom are significantly lower ($153.9$) than in the previous case. Also we have a slightly wider CI on this case.

In consequence, we must ask ourselves: is it reasonable to assume equal variances?
```{r}
#compute CI for sigma_1^2/sigma_2^2
var.test(x,y,conf.level = 0.99)
```
As the CI provided by this test includes the value 1 we can a assume (with a 99% confidence level) that the variances are equal.

## Variance CI

We can also program a user-costumed function to compute the Variance Confidence Interval.

```{r}
#customized function
NormalVar_CI <- function(x, lvl=0.95){
  
  #known parameters
  alpha = 1-lvl
  n = length(x)
  
  #---------list to be filled
  params <- list(1,2,3,4,5)
  param_names <- c("Computed values",
                   "Lower bound", "Upper bound",
                   paste0("Confidence Interval at ",lvl*100,"%"),
                   "CI width")
  names(params) <- param_names
  
  #computed values
  s2 = var(x) #sample variance
  chi1 = qchisq(1-alpha/2, n-1)
  chi2 = qchisq(alpha/2, n-1)

  comp = c(s2, chi1, chi2)
  names(comp) = c("Sample variance", "Chisqare 1-alpha/2 quantile", "Chisqare alpha/2 quantile")
  params[[1]] <- comp
  
  #lower bound
  L1 = ((n-1)*s2)/chi1
  params[[2]] <- L1
  
  #upper bound
  L2 = ((n-1)*s2)/chi2
  params[[3]] <- L2
  
  #Confidence interval 
  CI = c(L1, L2)
  params[[4]] <- CI
  
  #width
  w <- diff(CI)
  params[[5]] <- w
  
  return(params)
}
```

As usual, we generate a random sample of variance 3.103449 (equal to the example on lecture with $(n-1)s^2=90$) then we imagine we do not know this variance. We want to know if we can assume the variance is 3. So we compute our function.
```{r}
#generate a random sample
x = rnorm(30, 0, sd=sqrt(3.103449))

#eval
a <- NormalVar_CI(x)
a
```
As the produced CI contains the value 3, we can say that 3 is a reasonable variance for the population. 

Two estimate the ratio of two samples variances we can use the `var.test()` function. This is useful if I want to see if one sample is more uncertain than other. 

```{r}
# generate two random samples
x = rnorm(50)
y = rnorm(30,1)

#compute the test 
var.test(x,y) #default conf.level=0.95
```

As we can see 1 is included on the confidence interval for a level of 95% of confidence. thus, we cannot reject that the two variances are equal.

# Test hypothesis

## Test by the right

Say we have a population with gaussian distribution, and unknown variance. We want to test if the population mean is -2 or greater than 2 with a confidence level of 95%. 

We need to perform a t.test whose inputs are the observed sample, the mean population under the null hypothesis and the alternative hypothesis (less, greater, two.sided) and the confidence level.

```{r}
# H0: mu=-2 vs H1: mu > -2, alpha = 0.05

x <- rnorm(7, -2, 1) #observed sample

t.test(x, mu=-2, alternative = "greater", conf.level=0.95)
```
as $pvalue>0.05$ we do not reject the null hypothesis. If we want to reproduce some of this result, for example, the standardized test statistic or the p-value: 

```{r}
t = (mean(x)-(-2))/(sd(x)/sqrt(length(x)))
t

p = 1 - pt(t, 6)
p
```
## Bilateral test

Now we have a gaussian population with unknown variance. The null hypothesis states the population mean is 2 and the alternative any number different from 2. Do we reject with a level of confidence of 95%?

```{r}
t.test(x, mu=2, alternative = "two.sided", conf.level=0.95)
```
We reject since 1) $p value < 0.05$ but also 2) the 95% confidence interval does not contain the population mean proposed by the null hypothesis. Moreover, 3) the standardized statisctic is lower than the cat-off t: $t_{obs} < t_{95%} = qt(0.05,6) = -1.94$.

```{r totalclean, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
rm(list = ls())
gc()
```
